A Simple Minibatch Acceptance Test for MCMC

### 2016-05-18 nips_2016.tex
Substantially updated section 3. Now sections 1-3, maybe 4, look good, first 5-6 pages look passable

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \prod_{i=1}^n p(x_i \mid \theta') q(\theta_t \mid
\theta')}{p(\theta_t) \prod_{i=1}^n p(x_i \mid \theta_t) q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are the likelihood ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian as $n
\to \infty$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution ($\mu=0$, $s=1$); we use three standard deviation (SD) values of
    $X_{\rm norm}$. The red $X_{\rm norm}$ and $X_{\rm corr}$ curves convolve to form the logistic,
    etc. The $y$-axis is capped at 0.5 for readability. This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$.

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution\footnote{Computing the distributions of $\varepsilon$ and
$X_{\rm norm}$ reduces to finding their standard deviations.} of $\varepsilon$ (from
Equation~\ref{eq:relationship}), then we can pretend that our $\Delta'$ is in fact $\Delta$. To
``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its components, $X_{\rm
norm}$, plays the part of $\varepsilon$. In light of our revised MH test, the previous discussion
raises the question: in an iteration of MCMC with current $\theta_t$ and proposed $\theta'$, how do
we know the distribution of $X_{\rm corr}$? We need the correct distribution to sample $X_{\rm
corr}$ appropriately. This procedure consists of two major steps.

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}), $K$ is
    usually small, around 5-10, and $n$ is usually a few hundred for the Central Limit Theorem to
    apply.  Using these $K$ values, we use the empirical standard deviation ${\rm std}(\Delta')$.

@@

Figure~\ref{fig:deconvolution} demonstrates this deconvolution process and provides three examples
of densities for $X_{\rm norm}$ and $X_{\rm corr}$. We know that the densities of $X_{\rm norm}$ and
$X_{\rm corr}$ convolve to form the density of a standard logistic random variable with mean $\mu=0$
and scale $s=1$. If $X_{\rm norm}$ has small standard deviation, then the deconvolution to form
$X_{\rm corr}$ will result in a wider distribution. Likewise, with a larger standard deviation for
$X_{\rm norm}$, the $X_{\rm corr}$ distribution is narrower, but note that it becomes increasingly
unstable with bumps in the distribution, because the logistic curve has fatter tails than the
Gaussian curve. For this reason, we cap the standard deviation possibilities of $X_{\rm norm}$ at
$\approx 1.2$. We also set the minimum standard deviation of $X_{\rm norm}$ to be 0.1, because below
that, the $X_{\rm corr}$ distribution essentially matches the distribution of the logistic.

Algorithm~\ref{alg:our_algorithm} concisely describes our MH test.

Since our proposed acceptance function $g(\Delta) = (1+ \exp(-\beta \Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
approximation in Equation~\ref{eq:criteria} an equality.

It is also important to note that our minibatch MH test can also be used to obtain parameter
estimates close to the MAP and/or ML estimates by augmenting the likelihood function with a
temperature term $T > 1$. Thus, the target distribution would become
\begin{equation}
\label{eq:temperature} f(\theta') = p(\theta')\left ( \prod_{i=1}^N p(x_i \mid \theta')\right)^{\frac{1}{T}}.
\end{equation}
We can decrease the temperature as the MCMC sampling proceeds, thus slowing the mixing rate and
sharpening the peaks of the distribution, so that the samples converge to the desired point estimate.

\begin{algorithm}[t]
\KwData{$\{x_i\}$ under a target distribution}
\KwResult{A series of sampled $\{\theta_i\}$ under the posterior distribution} 
pre-compute the distribution of $X_{corr}$ under the different value of $var(\Delta')$\;
\While{not enough sampled data}{
	estimate the $sd(\Delta')$\;
	choose the distribution of $X_corr$ from the pre-computed distribution family\;
    propose a candidate $\theta'$\;
    compute the noisy $\Delta'$\;
    sample a $x$ from $X_{corr}$\;
    \eIf{$x + \Delta' > 0$}{
        Accept the $\theta'$;
    }{
        Reject the $\theta'$;
    }
}
\caption{A description of our MH test. {\color{blue} TODO}}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

{\color{blue}
Daniel: lots of things to do for this section, such as revising the Algorithm 1 description,
breaking this up into subsections (it's a bit long), and also just making sure it is clear for
readers to understand.
}

@@

In this section, we explore the convergence properties of our algorithm.  Define our estimate of
$X_{\rm norm}$ as $X_{\rm norm}'$, and let $X_{\xi} = X_{\rm norm} - X_{\rm norm}'$, so that
$\Delta' = \Delta + X_{\rm norm} + X_\xi$.  We also denote $F_X$ as the CDF of $X$.  Let the
transition kernel of complete data at sampling step $i$ be $T_i$, and let the transition kernel of
minibatch data at sampling step $i$ is $T_i'$.  We show that the distance of $T_i$ and $T_i'$ is
proportional to our estimate error $X_{\xi}$. We also define the stationary distribution obtained
by $T_i$ is $\pi$.  We show the by decreasing the step size, the distribution of our sampled data
converges to $\pi$. We use something called the \emph{total variation distance} between two kernels,
denoted as $\| \cdot \|$.

>>>>

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') (\prod_{i=1}^n p(x_i \mid
\theta'))^{\frac{N}{n}} q(\theta_t \mid \theta')}{p(\theta_t) (\prod_{i=1}^n p(x_i \mid
\theta_t))^{\frac{N}{n}} q(\theta' \mid \theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are the likelihood ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively, and that the latter uses an
$N/n$ scaling term. We assume $n \ll N$ (so $\Delta'$ is substantially faster to compute) and that
the $n$ data points $x_i$ represent a random minibatch of data.

@@

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\frac{N}{n}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian as $n
\to \infty$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution. We use three standard deviation values of $X_{\rm norm}$. The
    two red curves convolve to form the logistic, etc. The $y$-axis is capped at 0.5 for
    readability. This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$. If $X_{\rm norm}$
is exact, then we get the same MH test with $\Delta'$ as we do with the full data version $\Delta$.

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution\footnote{Computing the distributions of $\varepsilon$ and
$X_{\rm norm}$ reduces to finding their standard deviations.} of $\varepsilon$ (from
Equation~\ref{eq:relationship}), then we can pretend that our $\Delta'$ is in fact $\Delta$. To
``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its components, $X_{\rm
norm}$, plays the part of $\varepsilon$. In light of our revised MH test, the previous discussion
raises the question: in an iteration of MCMC with current $\theta_t$ and proposed $\theta'$, how do
we know the distribution of $X_{\rm corr}$? We need the correct distribution to sample the
\emph{values} of $X_{\rm corr}$ appropriately. This procedure consists of two major steps.

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}), $K$ is
    small, around 5-10, and $n$ is usually a few hundred for the Central Limit Theorem to apply.
    Using these $K$ values, we use the empirical standard deviation ${\rm std}(\Delta')$.

@@

Algorithm~\ref{alg:our_algorithm} describes our MH test within the MCMC algorithm.

\subsection{Discussions and Practical Considerations of the New MH Test}\label{ssec:discussion}

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, number of $\Delta'$ estimates $K$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distributions, $\{x_i\}_{i=1}^n$ values from a target distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or
point estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$ and create an empty list for $\Delta'$ estimates, $d = []$\;
    \For{$k = \{1, \ldots, K\}$}{
        Sample a random minibatch of size $m$ from the complete data\;
        Compute $\Delta'$ from this minibatch (see Equation~\ref{eq:deltas}), and add $\Delta'$ to $d$\;
    }
    Compute ${\rm std}(d)$, the standard deviation estimate of $\Delta'$ from list $d$\;
    \eIf{${\rm std}(d) \ge 1.2$}{
        Skip loop (set $\theta_{t+1}=\theta_t$) and apply variance preconditioning fix (if desired)\;
    }{
        Choose the closest $X_{\rm corr}$ distribution, and sample a value $X_c$ from it\;
        Using another random minibatch of $m$ data points, compute a \emph{new} $\Delta'$ (call it $\Delta_{\rm real}'$)\;
        \eIf{$X_c + \Delta_{\rm real}' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old one, $\theta_{t+1} = \theta_t$\;
        }
    }
}
\caption{A description of our MH test within the MCMC algorithm.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

\textbf{The Convolution}. Figure~\ref{fig:deconvolution} demonstrates
Equation~\ref{eq:deconvolution} and provides three (color-coded) examples of densities for $X_{\rm
norm}$ and $X_{\rm corr}$. The density of $X$, a logistic random variable with mean $\mu = 0$ and
scale $s=1$, is known, as are the $X_{\rm norm}$ densities for different standard deviations. The
deconvolution provides us with the $X_{\rm corr}$ distributions. Note, however, that as the standard
deviation of $X_{\rm norm}$ increases, the $X_{\rm corr}$ distribution becomes increasingly unstable
and ``bumpier'', because the logistic curve has fatter tails than Gaussians. For this reason, we cap
the standard deviation possibilities of $X_{\rm norm}$ at $\approx 1.2$ in our experiments. In our
experiments, we discretize the possible standard deviations of $X_{\rm norm}$, and we also limit the
densities to consider the range $[-10,10]$ (as shown in Figure~\ref{eq:deconvolution}).

\textbf{Variance Preconditioning}. Returning to the discussion from Section~\ref{sec:introduction},
our test requires a variance check. Equation~\ref{eq:deconvolution} means that the variance of $X$,
which is $\pi^2/3\approx 3.29$ is an upper bound on the variances of $X_{\rm norm}$. Therefore, the
test cannot work if the variance of $X_{\rm norm}$ is too large. Since the standard deviation of $X$
is about $\sqrt{3.29}\approx 1.81$, this bounds ${\rm std}(X_{\rm norm})$ (as well as ${\rm
std}(\Delta')$). From the discussion above, we bound the standard deviation at 1.2 so that $X_{\rm
corr}$ can handle the remaining noise. Recall from Section~\ref{ssec:deltas_minibatch} that we
estimate the ${\rm std}(X_{\rm norm})$ each iteration. If the variance/standard deviation test
fails, then we can (a) skip the iteration, (b) increase the minibatch size, (c) change the proposal
to decrease step sizes, or (d) increase temperature, as we now discuss.

\textbf{Temperature}. One way to satisfy the variance precondition is to increase the temperature of
the target distribution. For a temperature $T>1$, the augmented target distribution would become
\begin{equation}\label{eq:log_temperature}
\log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
with an extra $1/T$ term augmented. As the temperature increases, the values of $\Delta'$ get
smaller, thus reducing variance. The flatter posterior results in easier mixing of the samples, and
as the samples approach a mode, one can decrease the temperature to sharpen the posterior peaks.

@@

In this section, we explore the convergence properties of our algorithm. Let our true and estimated
acceptance probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm
corr} > 0)$. Define our estimate of $X_{\rm norm}$ as $X_{\rm norm}'$, and let $X_{\xi} = X_{\rm
norm} - X_{\rm norm}'$, so that $\Delta' = \Delta + X_{\rm norm} + X_\xi$.  We also denote $F_X$ as
the CDF of $X$.  Let the transition kernel of complete data at sampling step $i$ be $T_i$, and let
the transition kernel of minibatch data at sampling step $i$ is $T_i'$.  We show that the distance
of $T_i$ and $T_i'$ is proportional to our estimate error $X_{\xi}$. We also define the stationary
distribution obtained by $T_i$ is $\pi$.  We show the by decreasing the step size, the distribution
of our sampled data converges to $\pi$. We use something called the \emph{total variation distance}
between two kernels, denoted as $\| \cdot \|$.
